Peptide retention time prediction

### 2015-11-09 retention.tex
dirks edits

Today a majority of the mass spectrometry (MS) protocols in use have a
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms for data interpretation a major
area within proteomics research. A particular difficult challenge in
this context is posed by the tremendous complexity of the data.
Proteins display a wide range of physico-chemical properties and are
often found at very different amounts in complex samples, spanning as
much as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications~\cite{Lemeer2009}. To manage this
diversity, increasing efforts are made to extract besides the
fragmentation spectra additional information from each MS experiment.

@@

However, in practice such predictive models is not as straight forward
as it might seem. For instance, one important part of such a model
would be the peptides three dimensional conformation. Although the
current structure prediction methods are quite accurate in predicting
three dimensional conformation of short amino acid chains, such
methods will not be equipped with the flexibility to model the various
conditions of a chromatographic experiment. For instance, many
peptides are likely changing their confirmation as a function of the
amount of organic solvent in the mobile phase, an effect that is
likely to have an important contribution to each peptides retention
time. Hence, most efforts to predict retention time have moved into
treating the chromatographic process as a black box, that is a system
that is only viewed in terms of its inputs an outputs.

Following these observations,a variety of retention time prediction methods has been proposed. Most naturally one could divide these
methods into four classes: look-up tables-based approaches,
index-based methods, modeling-based methods and machine learning
techniques. The main ideas underlying each category are illustrated in
Figure \ref{fig:methods} , while Table \ref{tab:rtmethods} lists a few
of the most popular retention time prediction methods from each
category.

@@

A limitation of index-based methods are that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
Elude} \cite{elude1,elude2} estimates retention time indices for each
dataset as hand. In these cases the indices are used to generate
features that are used as an input for final retention time
estimates. We will come back to this in section \ref{sec:ML}.

@@

In other modeling-based approaches theoretical models are used to
capture the chemical behavior of peptides on the chromatography
column. One such method is BioLCCC, models the chromatography process
using a lattice based random walk model for the peptides, which are
modeled as chains of monomers. The model also includes an entropy and
energy compensation for the amount of peptides residing on the column 
and a model of the interaction energies between the peptide and the
stationary phase.

@@

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer {\em et al.} proposed a predictor based on support vector
machines (SVMs) \cite{klammer2007improving}. Their method required
fewer training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction accuracy.
Subsequently, a more elaborate SVM-based method, {\sc RTPredict}, that
 use a customized kernel function was proposed \cite{rtpredict,
rtpredictImproved}. This approach provided an algorithm that required
as little as 50 peptides for training.

@@

In a benchmark based on the merit of how large retention time window,
measured as a fraction of the gradient time was be needed to include
95\% of the peptides of a series of 14 runs, {\sc Elude} outperformed
{\sc SSRCalc} and {\sc RTPredict}, needing only on average 22\% as
compared to 28\% and 26\% respectively~\cite{elude1}. However, despite
all these efforts, it should be pointed out that none of the retention
time prediction methods currently available is able to provide a
prediction precision comparable to the small shifts observed when
comparing retention time of peptides from two technical replicates
(Figure \ref{fig:repr}). Recently, the lack of suitable
representations for secondary structure elements was indicated as one
of the main causes of prediction errors \cite{Reimer2012}.  Most
probably additional factors are involved, and further studies are
required to even understand the error sources in the process.

@@

To address this, in Moruz {\em et al.} \cite{elude2} extended the
first version of their algorithm {\sc ELUDE} to encompass modified
peptides as well. This was straightforward to achieve once the authors
observed that many of their predictive features of their method were
all functions of the peptides amino acid sequence and a retention
index. As such retention indices are estimated for each dataset at
hand, at the beginning of the training procedure, they can trivially
be extended beyond the standard amino acids. When the authors used but
the features based on their retention time index, they obtained a
negligible loss in accuracy compared to using all their original
features for unmodified peptides. Since this index was trained for
each dataset at hand, the algorithm became fully trainable for any
modified amino acid (Figure \ref{fig:elude}). The new version of {\sc
ELUDE} was evaluated on peptides containing artifactual modifications
introduced as part of an MS protocol called COmbined FRActional
DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. The results
showed that {\sc ELUDE} yielded equally accurate predictions for
modified and unmodified peptides \cite{elude2}.

@@

One way to determine a predictable separation techniques usefulness is
to estimate its ability to remove candidate peptides of any detected
feature, normally measured by a false positive ratio (FPR), also known
as the methods fall-out. The FPR is of course a function of the
selected thresholds, but if we select a treshold were say $\sim$95\%
of the known examples get correctly included, what FPR do we find at
such level?  Current retention time prediction algorithms have an FPR
of about $\frac{1}{5}$, meaning that it can remove about four out of
five candidate peptides for a given feature. This could be compared to
the FPR of about $\frac{1}{200}$ for predictions of high-resolution
iso-electric focusing experiments~\cite{branca2014hirief}. The
retention time prediction have to improve quite a bit to come to level
with the fall-out of isoelectric focusing, however, almost all mass
spectrometry-based proteomics use reversed phase chromatography, while
very few studies are using iso-electric focusing.

>>>>

Today a majority of the mass spectrometry (MS) protocols in use have a
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms for data interpretation a major
area within proteomics research. A particularly difficult challenge in
this context is posed by the tremendous complexity of the data.
Proteins display a wide range of physico-chemical properties and are
often found at very different amounts in complex samples, spanning as
much as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications~\cite{Lemeer2009}. To manage this
diversity, increasing efforts are made to extract additional
information besides the fragmentation spectra from each MS experiment.

@@

However, in practice such predictive models are not as straight
forward to design as it might seem. For instance, one important part
of such a model would be the peptides three dimensional
conformation. Although the current structure prediction methods are
quite accurate in predicting three dimensional conformation of short
amino acid chains, such methods will not be equipped with the
flexibility to model the various conditions of a chromatographic
experiment. For instance, many peptides are likely changing their
confirmation as a function of the amount of organic solvent in the
mobile phase, an effect that is likely to have an important
contribution to each peptides retention time. Hence, most efforts to
predict retention time have moved into treating the chromatographic
process as a black box, that is, a system that is only viewed in terms
of its inputs and outputs.

Following these observations, a variety of retention time prediction
methods have been proposed. Most naturally one could divide these
methods into four classes: look-up tables-based approaches,
index-based methods, modeling-based methods and machine learning
techniques. The main ideas underlying each category are illustrated in
Figure \ref{fig:methods} , while Table \ref{tab:rtmethods} lists a few
of the most popular retention time prediction methods from each
category.

@@

A limitation of index-based methods are that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
  Elude} \cite{elude1,elude2} estimate retention time indices for each
dataset as hand. In these cases the indices are used to generate
features that are used as an input for final retention time
estimates. We will come back to this in section \ref{sec:ML}.

@@

In other modeling-based approaches theoretical models are used to
capture the chemical behavior of peptides on the chromatography
column. One such method is BioLCCC, which models the chromatography
process using a lattice based random walk model for the peptides. The
method models peptides as chains of monomers. The model also includes
an entropy and energy compensation for the amount of peptides residing
on the column and a model of the interaction energies between the
peptide and the stationary phase.

@@

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer {\em et al.} proposed a predictor based on support vector
machines (SVMs) \cite{klammer2007improving}. Their method required
fewer training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction accuracy.
Subsequently, a more elaborate SVM-based method, {\sc RTPredict}, that
uses a customized kernel function was proposed \cite{rtpredict,
  rtpredictImproved}. This approach provided an algorithm that
required as little as 50 peptides for training.

@@

In a benchmark based on the merit of how large the retention time
window, measured as a fraction of the gradient time was needed to
include 95\% of the peptides of a series of 14 runs, {\sc Elude}
outperformed {\sc SSRCalc} and {\sc RTPredict}, needing only on
average 22\% as compared to 28\% and 26\%
respectively~\cite{elude1}. However, despite all these efforts, it
should be pointed out that none of the retention time prediction
methods currently available is able to provide a prediction precision
comparable to the small shifts observed when comparing retention time
of peptides from two technical replicates (Figure
\ref{fig:repr}). Recently, the lack of suitable representations for
secondary structure elements was indicated as one of the main causes
of prediction errors \cite{Reimer2012}.  Most probably additional
factors are involved, and further studies are required to even
understand the error sources in the process.

@@

To address this, in Moruz {\em et al.} \cite{elude2} extended the
first version of their algorithm {\sc ELUDE} to encompass modified
peptides as well. This was straightforward to achieve once the authors
observed that many of the predictive features of their method were all
functions of the peptide's amino acid sequence and a retention
index. As such retention indices are estimated for each dataset at
hand, at the beginning of the training procedure, they can trivially
be extended beyond the standard amino acids. When the authors used but
the features based on their retention time index, they obtained a
negligible loss in accuracy compared to using all their original
features for unmodified peptides. Since this index was trained for
each dataset at hand, the algorithm became fully trainable for any
modified amino acid (Figure \ref{fig:elude}). The new version of {\sc
  ELUDE} was evaluated on peptides containing artifactual
modifications introduced as part of an MS protocol called COmbined
FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. The
results showed that {\sc ELUDE} yielded equally accurate predictions
for modified and unmodified peptides \cite{elude2}.

@@

One way to determine a predictable separation techniques usefulness is
to estimate its ability to remove candidate peptides of any detected
feature. This can be measured by a false positive ratio (FPR), also
known as the method's fall-out. The FPR is of course a function of the
selected thresholds, but if we select a threshold where $\sim$95\% of
the known examples get correctly included, what FPR do we find at such
level?  Current retention time prediction algorithms have an FPR of
about $\frac{1}{5}$, meaning that it can remove about four out of five
candidate peptides for a given feature. This could be compared to the
FPR of about $\frac{1}{200}$ for predictions of high-resolution
iso-electric focusing experiments~\cite{branca2014hirief}. The
retention time prediction has to improve quite a bit to come to level
with the fall-out of isoelectric focusing, however, almost all mass
spectrometry-based proteomics use reversed phase chromatography, while
very few studies are using iso-electric focusing.
